---
title: "Descriptive Statistics - Exercices"
author: "João Silva"
date: "2025-11-10"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,warning = F,message = F}
#install.packages(c("tidyverse", "gmodels", "table1", "readxl", "haven","kableExtra"))
library(tidyverse)
library(gmodels)
library(table1)
library(readxl)
library(haven)
library(kableExtra)
```


# A. Develop data

The table of data develop contains information on the development at age
5 and 8 of 39 children with a particular condition. To measure the
child’s development a scale measuring the overall development is used.
This scale is divided into 5 subscales: motor development, social
development, hearing-language development, hand-eye and practical
performance. In addition to child development at age 5 and 8, it was
also noted whether or not children had anemia before age 5 and the
socioeconomic class of the child's household.

Na tabela de dados registamos as seguintes variáveis:\
• **id** – child identification number in the study\
• **general1**- overall coefficient of development scale at 5 years\
• **general2**- overall coefficient of development scale at 8 years\
• **motor1**- motor coefficient of development scale at 5 years\
• **motor2**- motor coefficient of development scale at 8 years\
• **social1**- social coefficient of development scale at 5 years\
• **social2**- social coefficient of development scale at 8 years\
• **earlang 1**- hearing-language coefficient of development scale at 5 years\
• **earlang 2**- hearing-language coefficient of development scale at 8 years\
• **eyehand 1**- hand-eye coefficient of development scale at 5 years\
• **eyehand 2**- hand-eye coefficient of development scale at 8 years\
• **perf 1**- performance coefficient of development scale at 5 years\
• **perf 2**- performance coefficient of development scale at 8 years\
• **group** – child with Anemia? (1-“Anemic”: with anemia before 5 years /2-“Non
Anemic”: without anemia before 5 years)\
• **soclass** – Socioeconomic class (1-“high”; /2-“Medium” /3-“Low”)\

Each question should have as answer ONE single table, each table should present the
summary measures of the variables studied for each of the compared groups.

## 1. Will there be differences between the children of the three socioeconomic class groups with regard to general development and the various developmental subscales at age 8?


```{r}
develop <- read.csv("develop.csv")
# mfrow to plot multiple histograms. 2 rows and 3 columns.
par(mfrow=c(2,3))
hist(develop$general2, main="General 8 years",xlab="")
hist(develop$motor2, main="Motor 8 years", xlab="") 
hist(develop$social2, main="Social 8 years", xlab="")
hist(develop$earlang2, main="Ear-Language 8 years", xlab="") 
hist(develop$eyehand2, main="Eye-Hand 8 years", xlab="")
hist(develop$perf2, main="Performance 8 years", xlab="")
```
```{r}
# Rename the soclass variable with labels
develop$soclass <- factor(develop$soclass, levels=3:1, labels=c("Low","Medium","High"))
# Rename the group variable with labels
develop$group <- factor(develop$group, levels=1:2, labels=c("Anemic", "Not Anemic"))
# Rename the columns of develop data frame
colnames(develop) <- c("id", "general1", "general2", "motor1", "motor2", "social1", "social2", "earlang1", "earlang2", "eyehand1", "eyehand2", "perf1", "perf2", "group", "soclass")
# Create table 1 to see differences between the children of the three socioeconomic class groups with regard to general development and the various developmental subscales at age 8
table1 (~ general2 + motor2 + social2 + earlang2 + eyehand2 + perf2 | soclass, data=develop)
```


## 2. Between 5 and 8 years was there a difference (rise or fall) in overall development and the various development subscales?

```{r}
# Create column with age (5 or 8 years)
develop_long <- develop %>% pivot_longer(cols = c(general1, general2, motor1, motor2, social1, social2, earlang1, earlang2, eyehand1, eyehand2, perf1, perf2),
                                         names_to = c(".value", "age"),
                                         names_pattern = "(.*)([12])")
# Rename age values
develop_long$age <- factor(develop_long$age, levels=c("1", "2"), labels=c("5 years", "8 years"))
# Rename the columns of develop_long data frame
colnames(develop_long) <- c("id", "group", "soclass", "age", "general", "motor", "social", "earlang", "eyehand", "perf")
# Create table 1 to see differences between 5 and 8 years in overall development and the various development subscales
table1 (~ general + motor + social + earlang + eyehand + perf | age, data=develop_long)
```

## 3. Will there be differences between children in both groups (with anemia before 5 or without anemia before 5) regarding the various developmental scale coefficients at 5 years?

```{r}
# Create table 1 to see differences between children in both groups (with anemia before 5 or without anemia before 5) regarding the various developmental scale coefficients at 5 years
table1 (~ general1 + motor1 + social1 + earlang1 + eyehand1 + perf1 | group, data=develop)
```

## 4. Will there be differences in the proportion of above / below average children on the 5-year old overall developmental scale between children in both groups (with anemia before 5 or without anemia before 5)? And at 8 years?

```{r}
# Create a new variable overall_average_5 with above/below average labels
develop <- develop %>% mutate(overall_average_5 = ifelse(general1<mean(general1), "below", "=orabove"))
# Create table to see differences in the proportion of above / below average children on the 5-year old overall developmental scale between children in both groups (with anemia before 5 or without anemia before 5). dnn = dimension names
CrossTable(develop$group, develop$overall_average_5, format="SPSS", prop.chisq=F, prop.c=F, prop.t=F, digits=1, dnn = c("Group", "Overall Average 5 years"))
# Create a new variable overall_average_8 with above/below average labels
develop <- develop %>% mutate(overall_average_8 = ifelse(general2<mean(general2), "below", "=orabove"))
# Create table to see differences in the proportion of above / below average children on the 8-year old overall developmental scale between children in both groups (with anemia before 8 or without anemia before 8)
CrossTable(develop$group, develop$overall_average_8, format="SPSS", prop.chisq=F, prop.c=F, prop.t=F, digits=1, dnn = c("Group", "Overall Average 8 years"))
```

# B. Nutricao data

This 58-record data table contains data for babies at birth and after nine months.

## 1. Graph the variable length at birth (comp)

```{r}
# Read nutricao.sav
nutricao <- read_sav("nutricao.sav")
# Graph the variable length at birth (comp)
hist(nutricao$comp, main="Length at Birth", xlab="Length (cm)")
```

## 2. Describe the variable length at birth (comp) using the most appropriate summary measures.

```{r}
# Describe the variable length at birth (comp) using the most appropriate summary measures
mean_length <- mean(nutricao$comp, na.rm = T)
sd_length <- sd(nutricao$comp, na.rm = T)
summary_length <- data.frame(Mean = mean_length, SD = sd_length)
summary_length
```

## 3. Graph the ferritin variable at 9 months (ferrit9)

```{r}
# Graph the ferritin variable at 9 months (ferrit9)
hist(nutricao$ferrit9, main="Ferritin at 9 months", xlab="Ferritin (ng/mL)")
```

## 4. Describe the ferritin variable at 9 months (ferrit9) using the most appropriate summary measures.

```{r}
# Describe the ferritin variable at 9 months (ferrit9) using the most appropriate summary measures
median_ferritin <- median(nutricao$ferrit9, na.rm = T)
IQR_ferritin <- IQR(nutricao$ferrit9, na.rm = T)
summary_ferritin <- data.frame(Median = median_ferritin, IQR = IQR_ferritin)
summary_ferritin
```

## 5. Describe in each table each of the categorical variables: national vaccination program (PNV) and graffar index (graffar)

```{r}
# Describe the national vaccination program variable (pnv)
table_pnv <- table(nutricao$pnv)
prop_table_pnv <- prop.table(table_pnv)
table_pnv_df <- data.frame(Count = as.vector(table_pnv), Proportion = as.vector(prop_table_pnv))
rownames(table_pnv_df) <- names(table_pnv)
table_pnv_df
# Describe the graffar index variable (graffar)
table_graffar <- table(nutricao$graffar)
prop_table_graffar <- prop.table(table_graffar)
table_graffar_df <- data.frame(Count = as.vector(table_graffar), Proportion = as.vector(prop_table_graffar))
rownames(table_graffar_df) <- names(table_graffar)
table_graffar_df
```

## 6. Describe in a contingency table the national vaccination program variable (pnv) as a function of the graffar index variable

```{r}
# Create a contingency table of national vaccination program (pnv) as a function of graffar index
CrossTable(nutricao$pnv, nutricao$graffar, format="SPSS", prop.chisq=F, digits=1, dnn = c("National Vaccination Program", "Graffar Index"))
```

## 7. Graphically and wisely represent each of the categorical variables: national vaccination program (pnv) and graffar index (graffar)

```{r}
# Graphically represent the national vaccination program variable (pnv)
par(mfrow=c(1,2))
pie(table(nutricao$pnv), main="National Vaccination Program (PNV)")
barplot(table(nutricao$graffar), main="Graffar Index", xlab="Graffar Categories", ylab="Frequency")
```

## 8. Recode the variable number months with breastfeeding (leitemat) to make it a dichotomous variable (1 = 'less than 9 months', 2 = '9 months')

```{r}
# Recode the variable number months with breastfeeding (leitemat) to make it a dichotomous variable
nutricao <- nutricao %>% mutate(leitemat_cat = ifelse(leitemat < 9, "less than 9 months", "9 months"))
# View the recoded variable
table(nutricao$leitemat_cat)
```

## 9. Graph the birth weight variable (pn) in male and female newborns (sex)

```{r}
# Graph the birth weight variable (pn) in male and female newborns (sexo)
boxplot(pn ~ sexo, data = nutricao, main="Birth Weight by Sex", xlab="Sex", ylab="Birth Weight (g)", names=c("Female", "Male"))
```

## 10. Graph the weight variable at 9 months (peso9) in babies taking vitamin C (vitc) and with the updated national vaccination plan (pnv)

```{r}
# Graph the weight variable at 9 months (peso9) in babies taking vitamin C (vitc) and with the updated national vaccination plan (pnv)
nutricao <- nutricao %>% mutate(vitc_label = ifelse(vitc == 1, "Taking", "Not Taking"),
                                pnv_label = ifelse(pnv == 1, "Updated", "Not Updated"))
par(mfrow=c(1,2))
boxplot(peso9 ~ vitc_label, data = nutricao, main="Weight at 9 months by Vitamin C", xlab="Vitamin C Status", ylab="Weight at 9 months (g)")
boxplot(peso9 ~ pnv_label, data = nutricao, main="Weight at 9 months by PNV", xlab="PNV Status", ylab="Weight at 9 months (g)")
```

## 11. Create and graph the weight gain (inc_weight) variable that corresponds to the weight gain from birth (pn) to 9 months (peso9)

```{r}
# Create the weight gain (inc_weight) variable
nutricao <- nutricao %>% mutate(inc_weight = peso9 - pn)
# Graph the weight gain (inc_weight) variable
hist(nutricao$inc_weight, main="Weight Gain from Birth to 9 months", xlab="Weight Gain (g)")
```

# C. Sepsis data

This table of data with 300 records contains data related to sepsis admissions in Portuguese ICUs. Variables:\
• Hospitalization Id\
• Sex of the hospitalized: 1-Female; 2-Male\
• Date of birth of the hospitalized\
• Date of entry into the ICU\
• Date of exit from the ICU\
• Date of exit from hospital
• Hospitalization Result: 1-Survivor; 2-Deceased in ICU; 3-Deceased in hospital\
• Age at ICU entry\
• Length of stay\

## 1. Transfer this database to R and describe (graphically and with appropriate summary measures) the age at ICU admission and the length of stay in the ICU.

```{r}
#Open the data uci_sepsis.xls
uci_sepsis <- read_excel("uci_sepsis.xls")
# Rename the columns
colnames(uci_sepsis) <- c("id_hospitalizacao", "sexo", "data_nascimento", "data_entrada_uci", "data_saida_uci","data_saida_hospital", "resultado_internamento")
# Convert date columns to Date type with lubridate as_date
uci_sepsis <- uci_sepsis %>%
  mutate(data_nascimento = dmy(data_nascimento),
         data_entrada_uci = dmy(data_entrada_uci),
         data_saida_uci = dmy(data_saida_uci),
         data_saida_hospital = dmy(data_saida_hospital))
# Create variables idade_entrada_uci and tempo_internamento_uci
uci_sepsis <- uci_sepsis %>%
  mutate(idade_entrada_uci = time_length(difftime(as.Date(data_entrada_uci), as.Date(data_nascimento)),"year"),
         tempo_internamento_uci = time_length(difftime(as.Date(data_saida_uci), as.Date(data_entrada_uci)),"days"))
#Graphically represent the age at ICU admission and the length of stay in the ICU
par(mfrow=c(1,2))
hist(uci_sepsis$idade_entrada_uci, main="Age at ICU admission",xlab="")
hist(uci_sepsis$tempo_internamento_uci, main="Length of stay in ICU", xlab="")
```        


```{r}
#Descrition of the age at ICU admission and the length of stay in the ICU with appropriate summary measures
#mean(uci_sepsis$idade_entrada_uci, na.rm = T)
#sd(uci_sepsis$idade_entrada_uci, na.rm = T)
#mean(uci_sepsis$tempo_internamento_uci, na.rm = T)
#sd(uci_sepsis$tempo_internamento_uci, na.rm = T)
# Table with the means and standard deviations of age at ICU admission and length of stay in ICU
table <- uci_sepsis %>% summarise (mean_age = mean(idade_entrada_uci, na.rm = T), sd_age = sd(idade_entrada_uci, na.rm = T),
                          mean_stay = median(tempo_internamento_uci, na.rm = T), sd_stay = IQR(tempo_internamento_uci, na.rm = T))
# Rename table columns
colnames(table) <- c("Mean Age at ICU Admission (years)", "SD Age at ICU Admission (years)", "Median Length of Stay in ICU (days)", "IQR Length of Stay in ICU (days)")
table
```

## 2. What is the proportion of deceased in the female patient group and in the male patient group?

```{r}
# Create a new variable sexo_label with labels for the sexo variable
uci_sepsis <- uci_sepsis %>% mutate(sexo_label = ifelse(sexo == 1, "Female", "Male"))
# Create a new variable resultado_label with labels for the resultado_internamento variable
uci_sepsis <- uci_sepsis %>% mutate(resultado_label = case_when(
  resultado_internamento == 1 ~ "Survivor",
  resultado_internamento == 2 ~ "Deceased in ICU",
  resultado_internamento == 3 ~ "Deceased in Hospital"
))
# Create a column with survivor and deceased labels
uci_sepsis <- uci_sepsis %>% mutate(survivor_deceased = ifelse(resultado_internamento == 1, "Survivor", "Deceased"))
# Create a contingency table with the proportions of deceased in the female and male patient groups
CrossTable(uci_sepsis$sexo_label, uci_sepsis$survivor_deceased, format="SPSS", prop.chisq=F, prop.c=F, prop.t=F, digits=1, dnn = c("Sex", ""))
```

