---
title: "Linear Regression"
author: "João Silva"
date: "2026-01-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. The file tas_tad.sav contains the age and values of the systolic (SBP) and diastolic blood
pressure (DBP) in a random sample of 300 Portuguese individuals.

```{r}
# Install packages if not already installed
if (!require(tidyverse)) install.packages("tidyverse")
# Load the necessary libraries
library(tidyverse)
# Load the data
tas_tad <- haven::read_sav("tas_tad.sav")
```

## a. Represent graphically the possible relation between age (variable idade) and SBP
(variable tas).

```{r}
ggplot(tas_tad, aes(x = idade, y = tas)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Relation between Age and Systolic Blood Pressure (SBP)",
       x = "Age (idade)",
       y = "Systolic Blood Pressure (SBP - tas)") +
  theme_minimal()
```


## b. Represent graphically the possible relation between SBP (variable tas) and DBP
(variable tad).

## c. What is the Pearson correlation coefficient (r) between age (variable idade) and SBP
(variable tas)?

## d. What is the Pearson correlation coefficient (r) between SBP (variable tas) and DBP
(variable tad)?

## e. What is the determination coefficient (R2) for the relation between age and SBP? What
is the meaning of this value?

##f. What is the determination coefficient (R2) for the relation between SBP and DBP? What
is the meaning of this value?

## g. Taking into account the relationship of the SBP (variable tas) with age (variable idade)
or with weight (variable peso), indicate which of the two variables - age or weight -
best explains the variability observed in the SBP.

# 2. The file alcohol.sav contains a set of variables relating to 454 newborns from three
hospitals, including data related to mother of newborns, to pregnancy, to exposures and
habits during pregnancy, to delivery and the newborn itself at the time of delivery.

## a. Build a scatterplot to represent the relationship between the weight of the newborn
at birth (variable birthwt) and the gestational age (variable gestlmp). Comment on this
chart.

## b. Build a model that allows you to perceive and predict the behavior of the birth weight
(variable birthwt) as a function of the gestational age (variable gestlmp).

## c. Write the function (equation of the line) associated with the constructed linear model.

## d. What is the value and respective 95% confidence interval of the ordinate at origin of
the model and what is the interpretation of this parameter?

## e. What is the value and respective 95% confidence interval of the model slope and what
is the interpretation of this parameter?

## f. What is the model's predicted value for the weight of a newborn child 38 weeks'
gestational age? Compare this value with the average newborn weights at 38 weeks of
gestational age.

## g. Check the value of the coefficient of determination (R2) for this model, interpret its
value and confirm that it is equal to the ratio between the sum of squares explained
by the regression model and the total sum of squares.

## h. What are the fundamental assumptions that underlie this simple linear regression
model and that make the calculated parameter estimates valid?

## i. Check the model's assumptions by analyzing the residuals. Check the normality of the
residuals distribution, through the analysis of the histogram of the residuals, and
homoscedasticity, through the analysis of standardized residuals distribution
depending on the standardized predicted values.

# 3. Consider the data table bmd_drug.sav, which consists of 10268 recordings related to body
mineral density (BMD) and other possibly associated variables, including height (variable
altura), weight (variable peso) and age (variable idade).

## a. Compute the body mass index (BMI).

## b. Categorize the BMI in the following classes: <25 kg/m2, 25-30 kg/m2, >30 kg/m2.

## c. Are there differences in the BMD (variable bmdtot) mean between the BMI categories?

## d. Are there differences in the BMD (variable bmdtot) between the BMI categories
adjusting for age?

## e. Is the loss of BMD associated with age equal in the different groups of BMI?

## f. Evaluate the final model.