---
title: "Reliability and agreement exercises"
author: "João Silva"
date: "2025-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reliability and agreement exercises

## Introduction

A table of data agree.sav contains information on the diagnosis of a particular condition (1=yes or 0=no) of 54 children and on the development measurement in the same children.

The diagnosis was made independently by five different doctors (variables: doctor1, doctor2, doctor3, doctor4 and doctor5) and the development score was assessed by the same psychologist using two different battery of tests (variables: test1 and test2).

1. Assess the inter-rater agreement and reliably among doctors.

2. Are the two development tests intersubstitutable?

## Install and load packages: foreignm obs-agree, psy, boot

```{r, results='hide', include= FALSE}
# CRAN packages list
cran_packages <- c("foreign", "psy", "boot","tidyverse")

# Install only CRAN packages that are not already installed
for (pkg in cran_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

# Check and install obs.agree if necessary
if (!requireNamespace("obs.agree", quietly = TRUE)) {
  install.packages("obs.agree_1.0.tar", repos = NULL, type = "source")
}

# Load all packages
all_packages <- c(cran_packages, "obs.agree")
lapply(all_packages, library, character.only = TRUE)
```


## Get data from SPSS file

```{r}
# Read the SPSS file named "agree.sav" using the read.spss function from the 'foreign' package
# The parameter use.missing=TRUE indicates that missing values in SPSS should be preserved
# After file=, you must enter the path to the agree.sav file on your computer.
agree <- read.spss(file="agree.sav", use.missing=TRUE)
# Convert the 'agree' object (which is a special 'data.frame' class from read.spss) into a standard R data.frame
# This ensures compatibility with other functions that expect a common data.frame
agree <- as.data.frame(agree)
# Convert the 'agree' data.frame into a data matrix (data.matrix)
# This transforms all columns into numeric types if possible, creating a numeric matrix
# Useful for analyses requiring matrix data, but may lose factor labels or strings
agree <- data.matrix(agree)
```

## Get data from SPSS file (from haven package in tidyverse)

```{r}
#agree <- haven::read_sav("agree.sav")
```


## Question 1: assess the inter-rater agreement and reliability among doctors

### To asssess the agreement: Proportions of overall and specific agreement 

```{r}
RAI(agree[,c(1,2,3,4,5)])
```


If one doctor, selected at random, makes a diagnosis (particular condition present: yes or no), the probability of another doctor making an equal diagnosis is 0.896

If one doctor, selected at random, makes the diagnosis yes (the particular condition is present), the probability of another doctor making the same diagnosis is 0.942

If one doctor, selected at random, makes the diagnosis no (the particular condition is not present), the probability of another doctor making the same diagnosis is 0.517

# To assess reliability: Kappa 

```{r}
lkappa(agree[,c(1,2,3,4,5)])
```
The reliability of raters was K=0.473.

```{r}
lkappa.boot <- function(data,x) {lkappa(data[x,])}
res <- boot(agree[,c(1,2,3,4,5)],lkappa.boot,1000)
boot.ci(res,type="bca")
```


## Question 2: to know if the tests are intersubstitutable you can do the Bland and Altaman limits of agreemnt (try it with JAMOVI :-) )

### It is also possible to assess the reliability of scores obtained by the two testes using the Intraclass Correlation coefficient, ranging from 0 (no reliability) to 1 (maximum reliability).

```{r}
icc(agree[,c(6,7)])
```

### It is also possible to assess the disagreement of scores obtained by the testes using the Information Based Measure of disagreement ranging from 0 (no disagreement=total agreement) to 1 (maximum disagreement).

```{r}
IBMD(agree[,c(6,7)])
```

# Bland and Altman

```{r}
devtools::install_github("deepankardatta/blandr")
library(blandr) 
library(ggplot2) 
blandr.statistics ( agree$test1 , agree$test2 , sig.level=0.95 ) 
blandr.draw ( agree$test1 , agree$test2 , plotter='rplot' )
```


