---
title: "Your First Machine Learning Project in R"
author: "Based on Jason Brownlee's Tutorial"
output: html_document
---

# 1. Install and Load Packages

First, we need to install and load the necessary packages. The tutorial relies heavily on the `caret` package. Note: The source comments suggest installing `ellipse` for specific plots [1].

```{r setup, message=FALSE, warning=FALSE}
# Install packages (Run these lines once in the console if not installed)
# install.packages("caret", dependencies=c("Depends", "Suggests"))
# install.packages("ellipse")
# Load the package
library(caret)
```

2. Load the Data
We will use the famous iris dataset available in R.

```{r setup, message=FALSE, warning=FALSE}
# Attach the iris dataset to the environment
data(iris)
# Rename the dataset
dataset <- iris
```

3. Create a Validation Dataset
We split the loaded dataset into two parts: 80% for training and 20% for validation to hold back for final testing.

```{r setup, message=FALSE, warning=FALSE}
# Create a list of 80% of the rows in the original dataset we can use for training
validation_index <- createDataPartition(dataset$Species, p=0.80, list=FALSE)

# Select 20% of the data for validation
validation <- dataset[-validation_index,]

# Use the remaining 80% of data to training and testing the models
dataset <- dataset[validation_index,]
```

4. Summarize Dataset
We inspect the dimensions, types, class levels, and statistical summary of the data.

```{r setup, message=FALSE, warning=FALSE}
# Dimensions of dataset
dim(dataset)
```


```{r setup, message=FALSE, warning=FALSE}
# List types for each attribute
sapply(dataset, class)
```


```{r setup, message=FALSE, warning=FALSE}
# Take a peek at the first 5 rows of the data
head(dataset)
```


```{r setup, message=FALSE, warning=FALSE}
# List the levels for the class
levels(dataset$Species)
```


```{r setup, message=FALSE, warning=FALSE}
# Summarize the class distribution
percentage <- prop.table(table(dataset$Species)) * 100
cbind(freq=table(dataset$Species), percentage=percentage)

# Summarize attribute distributions
summary(dataset)
```

5. Visualize Dataset
Univariate Plots
We split the input (x) and output (y) to create individual plots.

```{r setup, message=FALSE, warning=FALSE}
# Split input and output
x <- dataset[,1:4]
y <- dataset[,5]

# Boxplot for each attribute on one image
par(mfrow=c(1,4))
for(i in 1:4) {
  boxplot(x[,i], main=names(iris)[i])
}
```


```{r setup, message=FALSE, warning=FALSE}
# Reset layout
par(mfrow=c(1,1))

# Barplot for class breakdown
plot(y)
```

Multivariate Plots
We visualize the interactions between variables. Note: featurePlot with plot="ellipse" requires the ellipse package installed.

```{r setup, message=FALSE, warning=FALSE}
# Scatterplot matrix
featurePlot(x=x, y=y, plot="ellipse")
```


```{r setup, message=FALSE, warning=FALSE}
# Box and whisker plots for each attribute
featurePlot(x=x, y=y, plot="box")
```


```{r setup, message=FALSE, warning=FALSE}
# Density plots for each attribute by class value
scales <- list(x=list(relation="free"), y=list(relation="free"))
featurePlot(x=x, y=y, plot="density", scales=scales)
```

6. Evaluate Algorithms
We set up a test harness using 10-fold cross-validation and build 5 different models (LDA, CART, kNN, SVM, RF).

```{r setup, message=FALSE, warning=FALSE}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="cv", number=10)
metric <- "Accuracy"

# a) Linear algorithms
set.seed(7)
fit.lda <- train(Species~., data=dataset, method="lda", metric=metric, trControl=control)

# b) Nonlinear algorithms
# CART
set.seed(7)
fit.cart <- train(Species~., data=dataset, method="rpart", metric=metric, trControl=control)

# kNN
set.seed(7)
fit.knn <- train(Species~., data=dataset, method="knn", metric=metric, trControl=control)

# c) Advanced algorithms
# SVM
set.seed(7)
fit.svm <- train(Species~., data=dataset, method="svmRadial", metric=metric, trControl=control)

# Random Forest
set.seed(7)
fit.rf <- train(Species~., data=dataset, method="rf", metric=metric, trControl=control)
```

7. Select Best Model
We summarize the results and compare the accuracy of the models.

```{r setup, message=FALSE, warning=FALSE}
# Summarize accuracy of models
results <- resamples(list(lda=fit.lda, cart=fit.cart, knn=fit.knn, svm=fit.svm, rf=fit.rf))
summary(results)
```


```{r setup, message=FALSE, warning=FALSE}
# Compare accuracy of models
dotplot(results)
```


```{r setup, message=FALSE, warning=FALSE}
# Summarize Best Model (LDA)
print(fit.lda)
```

8. Make Predictions
Finally, we use the best model (LDA) to make predictions on the validation set we held back at the beginning.

```{r setup, message=FALSE, warning=FALSE}
# Estimate skill of LDA on the validation dataset
predictions <- predict(fit.lda, validation)
confusionMatrix(predictions, validation$Species)
```

### Explanation of the Steps

1.  **Installation and Loading:** The code begins by installing the `caret` package (and `ellipse` for specific plots) if they are not already on your system, as this is the primary library used for the project [1, 20, 21].
2.  **Data Loading:** The script loads the built-in Iris dataset and renames it to `dataset` for consistency [2].
3.  **Validation Split:** It uses `createDataPartition` to split the data into 80% for training and 20% for validation. This ensures we have unseen data to test the final model's accuracy [3].
4.  **Summary:** The code generates various statistical summaries, including dimensions, attribute types, and class distributions, to understand the data structure [4-9, 22].
5.  **Visualization:** It separates inputs (`x`) and outputs (`y`) to create boxplots and barplots. It also uses `featurePlot` to create scatterplot matrices and density plots to visualize relationships [10-14].
6.  **Model Training:** The script defines a 10-fold cross-validation control and trains 5 different algorithms: Linear Discriminant Analysis (LDA), Classification and Regression Trees (CART), k-Nearest Neighbors (kNN), Support Vector Machines (SVM), and Random Forest (RF) [15, 16].
7.  **Comparison:** It compares the models using `resamples` and visualizes the accuracy spread using `dotplot`. In this tutorial, LDA was identified as the most accurate model [17, 18, 23].
8.  **Prediction:** The final step runs the LDA model on the validation dataset and uses a confusion matrix to verify the accuracy on new data [19].